---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# cier

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/mdsteiner/cier/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/mdsteiner/cier/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/mdsteiner/cier/graph/badge.svg)](https://app.codecov.io/gh/mdsteiner/cier)
<!-- badges: end -->

`cier` detects careless and insufficient-effort responding (C/IER) in survey data
using **indirect, response-pattern indices**. It favours a small, validated set of
indices and an auditable, false-positive-rate-controlled flagging layer over a large
method count -- every number traces to a paper or a trusted package. The core
battery follows the indirect indices evaluated by Goldammer et al. (2024);
standalone sequence, timing, and attention-check indices complement it.

## The indices

| Index | Function | Construct family | Flags | Default cutoff |
|---|---|---|---|---|
| Longstring | `cier_longstring` | -- | long identical runs | half the items |
| Response variability (IRV) | `cier_irv` | -- | low variability | 5% percentile |
| Even-odd consistency | `cier_even_odd` | consistency | low consistency | 5% percentile |
| Personal reliability (PR/RPR) | `cier_personal_reliability` | consistency | low reliability | 5% percentile |
| Psychometric synonyms | `cier_psychsyn` | -- | low agreement | 5% percentile |
| Psychometric antonyms | `cier_psychant` | -- | high agreement | 5% percentile |
| Mahalanobis distance | `cier_mahalanobis` | -- | multivariate outlier | chi-square, *p* < .001 |
| Person-total correlation | `cier_person_total` | -- | low correlation | 5% percentile |
| Gnormed (person-fit) | `cier_gnormed` | person-fit | aberrant pattern | Monte-Carlo null, *p* = .05 |
| Ht (person-fit) | `cier_ht` | person-fit | aberrant pattern | 5% percentile |

`cier_gnormed` / `cier_ht` use the suggested `PerFit` / `mokken` packages and are
skipped with a recorded reason when those are not installed.

Five further indices are standalone functions -- `cier_screen()` does not run them
(yet): they read inputs the response matrix does not carry (timestamps, check keys)
or overlap with the battery in ways still being evaluated.

| Index | Function | Family | Flags | Default cutoff |
|---|---|---|---|---|
| Autocorrelation | `cier_autocorrelation` | sequence pattern | repetitive / periodic answers | 5% percentile |
| Laz.R (Markov predictability) | `cier_lazr` | sequence pattern | predictable transitions | 5% percentile |
| Total time | `cier_total_time` | timing | fast completion | 5% percentile (lower) |
| Page time | `cier_page_time` | timing | pages under 2 s/item | any rapid page |
| Attention checks | `cier_attention` | direct | failed checks | any failed check |

## Installation

`cier` is not yet on CRAN. Install the development version from GitHub -- with
`dependencies = TRUE` so the suggested `PerFit` / `mokken` (which back Gnormed and
Ht) come along:

```{r, eval = FALSE}
# pak is the current r-lib installer:
# install.packages("pak")
pak::pak("mdsteiner/cier", dependencies = TRUE)
```

## Quick start

`cier_screen()` runs the whole battery over one response set and prints a
transparent flag table. The four metadata-using indices need an `items` frame
aligned to the columns; the bundled `bfi_careless` makes one easy to build from the
column names:

```{r}
library(cier)

resp <- bfi_careless[, 1:44]
nm   <- names(resp)
items <- data.frame(
  scale         = sub("^v_BFI_([A-Za-z]+)[0-9].*$", "\\1", nm),
  reverse_keyed = grepl("_R$", nm),
  max           = 5L
)

cier_screen(resp, items,
  control = list(cier_personal_reliability = list(seed = 1),
                 cier_gnormed = list(seed = 1)))
```

## Simulating careless responding

Alongside detection, `cier` can **generate** survey data with planted careless
responding -- a known pattern, extent, and onset in a known share of respondents
-- for power analysis, method comparison, and recovery tests. `cier_simulate()`
is a data generator, not an index: it has no cutoff and never enters
`cier_screen()`, and its output is *not* evidence of real-world validity (anchor
accuracy claims to labelled data).

```{r}
sim <- cier_simulate(
  n          = 120,
  items      = data.frame(scale = rep(c("E", "A", "C"), each = 6L), max = 5L),
  prevalence = 0.2,
  n_checks   = 2L,
  seed       = 2026
)
sim
```

Each slot feeds a shipped function directly -- `cier_screen(sim$responses,
sim$items)`, `cier_total_time(sim$seconds)`, `cier_page_time(sim$page_seconds,
sim$items_per_page)`, and `cier_attention(sim$checks, sim$pass)` -- scored
against the planted `sim$truth`. See `vignette("simulation")` for a worked
study.

## How the flagging works

The detection signal lives in **multi-index agreement**, not any single index.
Even-odd, PR and RPR measure one construct, so they collapse to **one vote** -- never
weighted as independent evidence. There is no trained classifier in this release.

Three things to keep in mind:

- The percentile cutoff is a **ranking convention**, not a calibrated false-positive
  rate -- it flags roughly `fpr`% of *this* sample by construction (more when
  coarse-scale responses tie at the cutoff), and is not Goldammer's simulated-null
  Sen95 operating point. (Mahalanobis and Gnormed are the exceptions: they reference
  a calibrated null instead.)
- The bundled attention-check labels are a **partial, biased** signal, not ground
  truth -- design checks in advance and do not treat them as a key.
- **Report results before and after exclusion**, and show the flag rate across `fpr`
  (for example 0.01, 0.05, 0.10).

See `vignette("cier")` for the full walk-through.

## Licence

Package code is MIT-licensed. The bundled `bfi_careless` dataset is redistributed
under CC BY 4.0 (Brühlmann et al., 2020); see `LICENSE.note`.

## References

Brühlmann, F., Petralito, S., Aeschbach, L. F., & Opwis, K. (2020). The
quality of data collected online: An investigation of careless responding in a
crowdsourced sample. *Methods in Psychology*, 2, 100022.

Goldammer P., Stöckli, P. L., Escher, Y. A., Annen, H., Jonas, K., & 
Antonakis, J. (2024) Careless responding detection revisited: Accuracy of 
direct and indirect measures. *Behavior Research Methods*, 56, 8422-8449.
