---
title: "Getting started with cier"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with cier}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## What `cier` does

Careless and insufficient-effort responding (C/IER) is survey data from people who
were not engaging with the items. `cier` flags it with a lean set of published
indices; each one scores every respondent and flags a tail through a cutoff rule.

- **Response-pattern indices** -- longstring, intra-individual response variability
  (IRV), even-odd and personal reliability, psychometric synonyms and antonyms,
  Mahalanobis distance, person-total correlation, and two nonparametric person-fit
  statistics (Gnormed, Ht) -- the indirect battery evaluated by Goldammer et al.
  (2024) -- plus two sequence indices, autocorrelation (Gottfried et al., 2022)
  and Laz.R (Biemann et al., 2025).
- **Timing indices** -- total completion time and per-page speeding, for when the
  survey platform exported timestamps.
- **Direct checks** -- a failed-count over the attention-check items a study
  designed in.

`cier_screen()` runs the ten-index response-pattern battery in one call. The
sequence, timing, and direct indices are standalone functions for now -- the
sections below show each one -- pending an evaluation of what they add to the
screen's votes.

## The example data

`bfi_careless` is 394 crowdsourced respondents (Brühlmann et al., 2020) who answered
the 44-item Big Five Inventory alongside **two independent attention checks**.

```{r}
library(cier)
dim(bfi_careless)
names(bfi_careless)[c(1:3, 45:46)]
```

The first 44 columns are the BFI items; the last two -- `v_Bogus_Item` and `v_IRI`
-- are the attention checks. Those labels are useful context, but treat them with
care: an attention check is a **partial and biased** signal. A respondent can be
careless and still answer the check correctly, and a careful but figurative reader
can fail one. They have to be designed into a study in advance, and they are not
ground truth -- so this vignette does not score the indices against them.

## One index in isolation

Every index is a function on the response matrix. Resampled personal reliability
(RPR) is one of the stronger indirect indicators: it splits each scale in half many
times and correlates the halves, so a respondent whose answers are internally
inconsistent scores high. The consistency indices need some item metadata:
which scale each item belongs to, whether it is reverse-keyed, and the response
range (`max`, the largest response option, plus `min` when the scale is not
1-based). Build that from the BFI column names: the scale is the letters after
`v_BFI_`, a trailing `_R` marks a reverse-keyed item, and every item is coded
1..5. (One caveat: the dataset authors' own analysis treated one item, `v_BFI_OP9_R`, as
keyed differently than its name suggests and corrected it; keying by the item names,
as here, makes the consistency indices differ slightly from the paper's counts --
`vignette("published-results")` documents the difference.)

```{r}
resp <- bfi_careless[, 1:44]
nm   <- names(resp)
items <- data.frame(
  scale         = sub("^v_BFI_([A-Za-z]+)[0-9].*$", "\\1", nm),
  reverse_keyed = grepl("_R$", nm),
  max           = 5L
)

rpr <- cier_personal_reliability(resp, items, seed = 1)
rpr
```

The print-out names the **direction** (an inconsistent respondent scores high, so
the upper tail flags), the resolved **cutoff**, and the flag rate. Note the
**abstentions**: these are respondents `cier` could not score, returned as `NA`.
Here they are straightliners -- people who gave the same answer to every item, so
there is no within-person variation to correlate. (`seed` makes the random
split-halves reproducible.)

The scores and flags live in `$value` / `$flagged`, or as a tidy frame:

```{r}
head(as.data.frame(rpr))
```

## The whole battery

`cier_screen()` runs the screenable indices at once and prints a transparent flag
table. It reuses the same `items`:

```{r}
screen <- cier_screen(resp, items,
  control = list(cier_personal_reliability = list(seed = 1),
                 cier_gnormed = list(seed = 1)))
screen
```

Note that even-odd and personal reliability measure the same construct (internal 
consistency), so they collapse to one vote. Counting them twice would double-weight 
one signal. That is why ten indices become nine votes. 

`cier_psychsyn` / `cier_psychant` abstain (`0 / 0`) because the BFI's
inter-item correlations never exceed the default pairing threshold, so there are no
synonym or antonym pairs to score. `cier_psychsyn_critval()` and 
`cier_synonym_pairs()` can be used to investigate the item-pair correlations. 

The agreement block is the real signal: far more respondents are flagged by two or
more votes than independent indices would produce, and that excess concentration is
what points to genuine careless responding rather than index-specific noise.

`Skipped: 0` here because `PerFit` and `mokken` (which back Gnormed and Ht) are
installed; without them those two indices are skipped with a recorded reason in
`screen$skipped`, and the rest of the screen still runs.

Because the signal is in agreement, extract respondents by the number of constructs
that flagged them, then subset for downstream work:

```{r}
flagged <- cier_flagged_cases(screen, min_votes = 2)
length(flagged)
head(as.data.frame(screen))
```

## Surveys that mix response formats

The BFI above is the easy case: every item shares one 5-point format. Real
surveys often mix 4-, 5-, 6-, and 7-point scales, and the battery handles that
unevenly:

- **Gnormed is the one hard limit.** Its `PerFit` backend scores a single
  number of response options across all items (Niessen et al., 2016 had to
  recode their items to a common 4-point scale for the same reason), so on
  mixed-format data `cier_screen()` records `cier_gnormed` as skipped with that
  reason and the rest of the battery still runs. To use Gnormed anyway, score a
  same-format subset of items in a direct call.
- **Every other index runs**, but read the results with care: the published
  evidence for these indices comes almost entirely from uniform-format batteries.
  Curran (2016), for example, describes even-odd consistency "across a series of
  similar scales using the same response format", and `mokken` (behind `cier_ht`)
  warns when items have different numbers of response categories. How well the
  indices separate careless from careful respondents on mixed formats is largely
  untested in the literature.
- **Straightliners change shape on mixed formats.** A respondent who always
  picks the same option *position* produces varying raw values once the scale
  range changes between blocks, so the consistency indices score them (as
  highly consistent) instead of abstaining. `cier_longstring` and `cier_irv`
  still catch them -- another reason the screen runs the indices as a battery.
- **The standalone sequence indices want same-format blocks.** Gottfried et
  al. (2022) state for `cier_autocorrelation` that "only questions with the
  same answer scales should be analyzed at one time"; on a mixed survey, score
  each same-format block in its own call.

## Sequence patterns: autocorrelation and Laz.R

Two further pattern indices read the response *sequence* -- the answers in the
order they were clicked -- and target the repetitive or periodic profiles the
consistency family abstains on (a perfect straightliner has no within-person
variance to correlate, but scores maximally on both of these):

- `cier_autocorrelation()` (Gottfried et al., 2022) takes the largest absolute
  autocorrelation of each respondent's response vector across a range of lags.
  The default lag range follows the reference implementation
  (`max_lag = ncol - 3`); on a long battery that includes short high-lag
  slices whose correlation saturates near 1, so follow the paper's advice and
  restrict to low lags.
- `cier_lazr()` (Biemann et al., 2025) scores first-order Markov response
  predictability: how well each answer is predicted by the one before it. It
  generalises longstring -- straightlining, alternating ("seesaw"), and
  diagonal-lining all score high. `kneedle = TRUE` swaps the percentile cutoff
  for the paper's sample-specific elbow rule.

```{r}
cier_autocorrelation(resp, max_lag = 8L)
cier_lazr(resp)
```

Both indices read the as-clicked sequence, so they assume the items share one
response scale and are scored in administration order (for mixed formats, see
the previous section). And both are *pattern specialists*: Laz.R is by
construction blind to random responding (Biemann et al., 2025), and a random
careless profile is exactly the irregular kind of response vector these
indices cannot separate from careful responding. Treat them as coverage for
the repetitive profile, alongside longstring and IRV -- not as stand-alone
detectors.

## Timing: total time and page time

When the survey platform exports timestamps, response time is the cheapest
C/IER evidence there is, and on *fully* careless respondents one of the
strongest (Goldammer et al., 2024). The timing indices take the platform's
exports directly -- there is no response matrix:

```{r}
secs <- c(420, 380, 95, 510, 47, 460, 405, 88, 390, 430)
cier_total_time(secs)
```

`cier_total_time()` flags the fast tail (the lower-percentile twin of the
ranking convention below). One published alternative is anchored to the sample
median instead: `frac_median = 0.5` flags respondents faster than half the
median (Leiner, 2019) -- a cutoff that stays meaningful even when a large
share of the sample speeds, which is exactly where a percentile is least
defensible.

`cier_page_time()` catches within-survey speeding bursts that a plausible
total can hide: it counts each respondent's pages answered faster than
`min_seconds = 2` seconds per item (Bowling et al., 2023), and by default any
rapid page flags.

```{r}
page_seconds <- matrix(c(20, 18, 22,
                         19, 21, 20,
                          3,  2, 24,
                         22, 20, 19,
                          2,  2,  2),
                       nrow = 5L, byrow = TRUE)
cier_page_time(page_seconds, items_per_page = c(2, 2, 2))
```

Note what that default is: an **absolute, cited heuristic** -- deliberately
unlike the percentile indices, because plausible reading speed does not adapt
to a sample's contamination. It is a *floor* on effort, not a content check:
Biemann et al. (2025) report that a large share of whole-survey straightliners
pass a 2 s/item screen. Use timing alongside content indices, never alone.

## Direct checks: scoring attention-check items

The bundled data's two checks can be scored as an index instead of by hand.
`pass` lists, per check column, the response value(s) that count as passing --
here the bogus item passes on the non-endorsing options 1 and 2, and the
instructed item passes on the directed answer 0 (Brühlmann et al., 2020):

```{r}
cier_attention(bfi_careless[, c("v_Bogus_Item", "v_IRI")],
               pass = list(c(1, 2), 0))
```

Per check, the failures reproduce the paper's own counts -- 92 on the bogus
item, 96 on the instructed item; `vignette("published-results")` shows that
reproduction. The caveat from the top of this vignette applies with full
force: a failed check is a **partial and biased** signal, not ground truth,
and checks must be designed into the study in advance (Meade & Craig, 2012).
Used that way -- as one piece of converging evidence among several -- a direct
check is the cheapest anchor a screen can have. An unanswered check
contributes no evidence: cier counts failures among the checks a respondent
answered, and a respondent who answered none abstains.

## Choosing a cutoff is a ranking decision, not a calibration

Every percentile index cuts the tail at a target fraction (`fpr`, default 0.05). It
is essential to be clear on what that does **not** mean: it is a **ranking
convention**, not a calibrated false-positive rate. A sample percentile flags about
that fraction *by construction* -- exactly so for a continuous score, approximately
when scores tie (common with coarse Likert items):

```{r}
for (f in c(0.01, 0.05, 0.10)) {
  fl <- cier_irv(resp, fpr = f)$flagged
  cat(sprintf("fpr %.2f  ->  %.1f%% flagged\n", f, 100 * mean(fl, na.rm = TRUE)))
}
```

The rate tracks `fpr` because you *defined* it to. This is *not* Goldammer et
al.'s Sen95 operating point, which is estimated against a simulated null of honest
responding; `cier`'s percentile makes no claim about the true false-positive rate.
Use it to rank respondents and cut a tail of the size you choose.
(Mahalanobis distance and Gnormed are the exceptions: they reference a calibrated
null -- a chi-square tail and a Monte-Carlo null -- rather than a sample percentile.)

## Reporting

Because the cutoff is a ranking choice, the exclusion is a researcher decision, not a
fact the data hands you. Two habits keep that honest:

- **Report results before and after exclusion.** If a conclusion only holds once
  flagged respondents are removed, that is itself a finding to disclose.
- **Show the flag rate across `fpr`** (for example 0.01, 0.05, 0.10, as above) so
  readers can see how sensitive the exclusion is to a threshold you picked.

## References

Biemann, T., Koch-Bayram, I. F., Meier-Barthold, M., & Aguinis, H. (2025). Using
Markov chains to detect careless responding in survey research. *Organizational
Research Methods, 28*(4), 543-568.

Bowling, N. A., Huang, J. L., Brower, C. K., & Bragg, C. B. (2023). The quick and
the careless: The construct validity of page time as a measure of insufficient
effort responding to surveys. *Organizational Research Methods, 26*(2), 323-352.

Brühlmann, F., Petralito, S., Aeschbach, L. F., & Opwis, K. (2020). The quality of
data collected online: An investigation of careless responding in a crowdsourced
sample. *Methods in Psychology, 2*, 100022.

Curran, P. G. (2016). Methods for the detection of carelessly invalid responses in
survey data. *Journal of Experimental Social Psychology, 66*, 4-19.

Goldammer, P., Stöckli, P. L., Escher, Y. A., Annen, H., Jonas, K., & Antonakis, J.
(2024). Careless responding detection revisited: Accuracy of direct and indirect
measures. *Behavior Research Methods, 56*(8), 8422-8449.

Gottfried, J., Ježek, S., Králová, M., & Řiháček, T. (2022). Autocorrelation screening: A
potentially efficient method for detecting repetitive response patterns in
questionnaire data. *Practical Assessment, Research & Evaluation, 27*, Article 2.

Leiner, D. J. (2019). Too fast, too straight, too weird: Non-reactive indicators
for meaningless data in internet surveys. *Survey Research Methods, 13*(3),
229-248.

Meade, A. W., & Craig, S. B. (2012). Identifying careless responses in survey
data. *Psychological Methods, 17*(3), 437-455.

Niessen, A. S. M., Meijer, R. R., & Tendeiro, J. N. (2016). Detecting careless
respondents in web-based questionnaires: Which method to use? *Journal of Research
in Personality, 63*, 1-11.
