---
title: "Getting started with cier"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with cier}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## What `cier` does

Careless and insufficient-effort responding (C/IER) is survey data from people who
were not engaging with the items. `cier` flags it with **indirect, response-pattern
indices** -- longstring, intra-individual response variability (IRV), even-odd and
personal reliability, psychometric synonyms and antonyms, Mahalanobis distance,
person-total correlation, and two nonparametric person-fit statistics (Gnormed, Ht).
Each index scores every respondent and flags a tail through a cutoff rule. 
The current set follows the recommended indirect battery evaluated by Goldammer et al. 
(2024); more criteria may be added in later versions. T

## The example data

`bfi_careless` is 394 crowdsourced respondents (Brühlmann et al., 2020) who answered
the 44-item Big Five Inventory alongside **two independent attention checks**.

```{r}
library(cier)
dim(bfi_careless)
names(bfi_careless)[c(1:3, 45:46)]
```

The first 44 columns are the BFI items; the last two -- `v_Bogus_Item` and `v_IRI`
-- are the attention checks. Those labels are useful context, but treat them with
care: an attention check is a **partial and biased** signal. A respondent can be
careless and still answer the check correctly, and a careful but figurative reader
can fail one. They have to be designed into a study in advance, and they are not
ground truth -- so this vignette does not score the indices against them.

## One index in isolation

Every index is a function on the response matrix. Resampled personal reliability
(RPR) is one of the stronger indirect indicators: it splits each scale in half many
times and correlates the halves, so a respondent whose answers are internally
inconsistent scores high. The consistency indices need some item metadata:
which scale each item belongs to, whether it is reverse-keyed, and the response
range (`max`, the largest response option, plus `min` when the scale is not
1-based). Build that from the BFI column names: the scale is the letters after
`v_BFI_`, a trailing `_R` marks a reverse-keyed item, and every item is coded
1..5. (One caveat: the dataset authors' own analysis treated one item, `v_BFI_OP9_R`, as
keyed differently than its name suggests and corrected it; keying by the item names,
as here, makes the consistency indices differ slightly from the paper's counts --
`vignette("published-results")` documents the difference.)

```{r}
resp <- bfi_careless[, 1:44]
nm   <- names(resp)
items <- data.frame(
  scale         = sub("^v_BFI_([A-Za-z]+)[0-9].*$", "\\1", nm),
  reverse_keyed = grepl("_R$", nm),
  max           = 5L
)

rpr <- cier_personal_reliability(resp, items, seed = 1)
rpr
```

The print-out names the **direction** (an inconsistent respondent scores high, so
the upper tail flags), the resolved **cutoff**, and the flag rate. Note the
**abstentions**: these are respondents `cier` could not score, returned as `NA`.
Here they are straightliners -- people who gave the same answer to every item, so
there is no within-person variation to correlate. (`seed` makes the random
split-halves reproducible.)

The scores and flags live in `$value` / `$flagged`, or as a tidy frame:

```{r}
head(as.data.frame(rpr))
```

## The whole battery

`cier_screen()` runs the screenable indices at once and prints a transparent flag
table. It reuses the same `items`:

```{r}
screen <- cier_screen(resp, items,
  control = list(cier_personal_reliability = list(seed = 1),
                 cier_gnormed = list(seed = 1)))
screen
```

Note that even-odd and personal reliability measure the same construct (internal 
consistency), so they collapse to one vote. Counting them twice would double-weight 
one signal. That is why ten indices become nine votes. 

`cier_psychsyn` / `cier_psychant` abstain (`0 / 0`) because the BFI's
inter-item correlations never exceed the default pairing threshold, so there are no
synonym or antonym pairs to score. `cier_psychsyn_critval()` and 
`cier_synonym_pairs()` can be used to investigate the item-pair correlations. 

The agreement block is the real signal: far more respondents are flagged by two or
more votes than independent indices would produce, and that excess concentration is
what points to genuine careless responding rather than index-specific noise.

`Skipped: 0` here because `PerFit` and `mokken` (which back Gnormed and Ht) are
installed; without them those two indices are skipped with a recorded reason in
`screen$skipped`, and the rest of the screen still runs.

Because the signal is in agreement, extract respondents by the number of constructs
that flagged them, then subset for downstream work:

```{r}
flagged <- cier_flagged_cases(screen, min_votes = 2)
length(flagged)
head(as.data.frame(screen))
```

## Surveys that mix response formats

The BFI above is the easy case: every item shares one 5-point format. Real
surveys often mix 4-, 5-, 6-, and 7-point scales, and the battery handles that
unevenly:

- **Gnormed is the one hard limit.** Its `PerFit` backend scores a single
  number of response options across all items (Niessen et al., 2016 had to
  recode their items to a common 4-point scale for the same reason), so on
  mixed-format data `cier_screen()` records `cier_gnormed` as skipped with that
  reason and the rest of the battery still runs. To use Gnormed anyway, score a
  same-format subset of items in a direct call.
- **Every other index runs**, but read the results with care: the published
  evidence for these indices comes almost entirely from uniform-format batteries.
  Curran (2016), for example, describes even-odd consistency "across a series of
  similar scales using the same response format", and `mokken` (behind `cier_ht`)
  warns when items have different numbers of response categories. How well the
  indices separate careless from careful respondents on mixed formats is largely
  untested in the literature.
- **Straightliners change shape on mixed formats.** A respondent who always
  picks the same option *position* produces varying raw values once the scale
  range changes between blocks, so the consistency indices score them (as
  highly consistent) instead of abstaining. `cier_longstring` and `cier_irv`
  still catch them -- another reason the screen runs the indices as a battery.
- **The standalone sequence indices want same-format blocks.** Gottfried et
  al. (2022) state for `cier_autocorrelation` that "only questions with the
  same answer scales should be analyzed at one time"; on a mixed survey, score
  each same-format block in its own call.

## Choosing a cutoff is a ranking decision, not a calibration

Every percentile index cuts the tail at a target fraction (`fpr`, default 0.05). It
is essential to be clear on what that does **not** mean: it is a **ranking
convention**, not a calibrated false-positive rate. A sample percentile flags about
that fraction *by construction* -- exactly so for a continuous score, approximately
when scores tie (common with coarse Likert items):

```{r}
for (f in c(0.01, 0.05, 0.10)) {
  fl <- cier_irv(resp, fpr = f)$flagged
  cat(sprintf("fpr %.2f  ->  %.1f%% flagged\n", f, 100 * mean(fl, na.rm = TRUE)))
}
```

The rate tracks `fpr` because you *defined* it to. This is *not* Goldammer et
al.'s Sen95 operating point, which is estimated against a simulated null of honest
responding; `cier`'s percentile makes no claim about the true false-positive rate.
Use it to rank respondents and cut a tail of the size you choose.
(Mahalanobis distance and Gnormed are the exceptions: they reference a calibrated
null -- a chi-square tail and a Monte-Carlo null -- rather than a sample percentile.)

## Reporting

Because the cutoff is a ranking choice, the exclusion is a researcher decision, not a
fact the data hands you. Two habits keep that honest:

- **Report results before and after exclusion.** If a conclusion only holds once
  flagged respondents are removed, that is itself a finding to disclose.
- **Show the flag rate across `fpr`** (for example 0.01, 0.05, 0.10, as above) so
  readers can see how sensitive the exclusion is to a threshold you picked.

## References

Brühlmann, F., Petralito, S., Aeschbach, L. F., & Opwis, K. (2020). The quality of
data collected online: An investigation of careless responding in a crowdsourced
sample. *Methods in Psychology, 2*, 100022.

Curran, P. G. (2016). Methods for the detection of carelessly invalid responses in
survey data. *Journal of Experimental Social Psychology, 66*, 4-19.

Goldammer, P., Stöckli, P. L., Escher, Y. A., Annen, H., Jonas, K., & Antonakis, J.
(2024). Careless responding detection revisited: Accuracy of direct and indirect
measures. *Behavior Research Methods, 56*(8), 8422-8449.

Gottfried, J., Ježek, S., & Králová, M. (2022). Autocorrelation screening: A
potentially efficient method for detecting repetitive response patterns in
questionnaire data. *Practical Assessment, Research & Evaluation, 27*, Article 2.

Niessen, A. S. M., Meijer, R. R., & Tendeiro, J. N. (2016). Detecting careless
respondents in web-based questionnaires: Which method to use? *Journal of Research
in Personality, 63*, 1-11.
