---
title: "Simulation studies with cier"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation studies with cier}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(cier)
```

## What `cier_simulate()` is for -- and what it is not

`cier_simulate()` plants known careless responding into a survey dataset: it
generates attentive responses from a graded response model, then overwrites a
known share of respondents with a known pattern, extent, and onset, and adds
matching response times and optional attention checks. Because the truth is
planted, you can score any index against it. That makes the simulator useful for

1. **Power analysis and study design** -- given your `n`, scales, and expected
   contamination, what sensitivity can you expect?
2. **Method comparison on planted patterns** -- which index catches which kind
   of careless responder, on data you control?
3. **Package recovery and property tests** -- "plant pattern X, check that the
   matched index ranks it to the top".
4. **Replicating published simulation designs.**

It is **not** evidence of real-world validity. Simulated carelessness is more
stylized than even instructed carelessness, and the warning is concrete:
Schroeders et al. (2022) trained a detector on simulated data and found its
advantage did **not** transfer to real respondents. So treat every number in
this vignette as a statement about *this synthetic data*, and anchor any
real-world accuracy claim to labelled real data instead (see
`vignette("published-results")`). No simulated number belongs in a headline
accuracy table.

## A simulated dataset

The generator needs an `items` frame describing the battery: which `scale` each
item loads on (this defines the factor structure), whether it is
`reverse_keyed`, and the response range (`max`, plus `min` when the scale is not
1-based). Here are three eight-item scales on a 5-point format, every other item
reverse-keyed:

```{r}
items <- data.frame(
  scale         = rep(c("E", "A", "C"), each = 8L),
  reverse_keyed = rep(c(FALSE, TRUE), times = 12L),
  max           = 5L
)

sim <- cier_simulate(
  n          = 200L,
  items      = items,
  prevalence = 0.2,
  n_checks   = 2L,
  seed       = 2026L
)
sim
```

The print-out is derived from the planted truth: a fifth of respondents are
careless, split across the eight recognized patterns (the default weights them
equally), all with `full` extent here. The role-boundary reminder is the last
line, by design.

The returned object is a plain list whose slots each feed a shipped function
directly:

```{r}
str(sim, max.level = 1, give.attr = FALSE)
```

`$truth` carries one row per respondent -- the flag, the pattern (`"attentive"`
on clean rows), the extent, the careless span (`onset_item` / `offset_item`),
and the per-row generator knobs in `$params`:

```{r}
head(sim$truth[, c("careless", "pattern", "extent", "onset_item",
                   "offset_item", "speeded")], 6)
```

### Tuning a pattern

Each careless style has its own knobs, passed through `pattern_params` and
recorded verbatim in `truth$params`. An unknown knob name is an error, so a typo
cannot silently plant a default. For example, hold every straightliner on the
maximum response and flip it occasionally:

```{r}
tuned <- cier_simulate(
  n = 50L, items = items, prevalence = 0.4,
  patterns       = c(straightline = 1),
  pattern_params = list(straightline = list(anchor = "value", value = 5L,
                                            switch_prob = 0.1)),
  seed = 7L
)
tuned$truth$params[tuned$truth$careless][[1]]   # the knobs the engine applied
```

## Scoring the content battery

`cier_screen()` runs the response-pattern battery over `sim$responses`, reusing
`sim$items` exactly as a real analysis would:

```{r}
screen <- cier_screen(sim$responses, sim$items,
  control = list(cier_personal_reliability = list(seed = 1),
                 cier_gnormed = list(seed = 1)))
screen
```

Two things to read here. First, even-odd and personal reliability measure one
construct, so the ten indices collapse to nine votes -- the agreement block and
the construct count run on the collapsed votes, never the raw per-index flags.
Second, `cier_psychsyn` / `cier_psychant` abstain (`0 / 0`): with default
loadings of 0.7, two items on the same scale correlate about 0.49, below the
0.60 pairing threshold, so there are no synonym or antonym pairs to score --
exactly as on a real battery whose scales are not near-redundant.

The signal lives in **agreement**, so extract respondents by how many constructs
flagged them, then score that against the planted truth. Because the cutoff is a
ranking convention, treating *every* non-attentive pattern as "should exclude"
makes the sensitivity below a lower bound -- a `speeder` has attentive content
and is invisible to this battery on purpose (the next section catches it):

```{r}
truth <- sim$truth$careless
sens_spec <- function(flagged) {
  c(sensitivity = sum(truth & flagged) / max(sum(truth), 1L),
    specificity = sum(!truth & !flagged) / max(sum(!truth), 1L))
}

n <- nrow(sim$responses)
for (k in 1:3) {
  pred <- seq_len(n) %in% cier_flagged_cases(screen, min_votes = k)
  cat(sprintf("min_votes >= %d:  ", k))
  print(round(sens_spec(pred), 3))
}
```

Requiring more votes trades sensitivity for specificity -- the researcher's
threshold, made visible.

## Timing and attention checks

The timing and direct indices are standalone -- they read the slots
`cier_simulate()` produced, not the response matrix:

```{r}
cier_total_time(sim$seconds)
cier_page_time(sim$page_seconds, sim$items_per_page)
cier_attention(sim$checks, sim$pass)
```

Nothing here separates the groups perfectly: response times are drawn with a
per-respondent pace intercept so the attentive and careless distributions
overlap, and checks fail probabilistically on both sides. A rank-based AUC makes
that concrete. The package exports no AUC helper -- it is a one-liner, with a
direction argument because some indices flag *low* scores (fast total time)
and some flag *high* ones (failed checks). For a comparison table, keep the
direction map beside the methods you score:

```{r}
rank_auc <- function(value, careless, higher_is_careless = TRUE) {
  ok <- !is.na(value)
  v <- value[ok]; y <- careless[ok]
  if (!higher_is_careless) v <- -v
  n1 <- sum(y); n0 <- sum(!y)
  if (n1 == 0L || n0 == 0L) return(NA_real_)
  (sum(rank(v)[y]) - n1 * (n1 + 1) / 2) / (n1 * n0)
}

flag_direction <- c(
  cier_longstring = "upper",
  cier_autocorrelation = "upper",
  cier_mahalanobis = "upper",
  cier_lazr = "upper",
  cier_total_time = "lower",
  cier_attention = "upper"
)

# TRUE when HIGH scores flag carelessness.
flag_is_upper <- function(method) {
  identical(flag_direction[[method]], "upper")
}

c(total_time = rank_auc(cier_total_time(sim$seconds)$value, truth,
                        flag_is_upper("cier_total_time")),
  attention  = rank_auc(cier_attention(sim$checks, sim$pass)$value, truth,
                        flag_is_upper("cier_attention")))
```

Total time lands inside the .66--.92 range reported across real datasets
(Goldammer et al., 2024), never at 1.0 -- a simulator on which everything scores
perfectly would be wrong by construction.

### The speeder split

A `speeder` is the case timing exists for: attentive *content*, careless
*times*. The content battery should miss it, and timing should catch it. Among
the planted speeders:

```{r}
speeder  <- sim$truth$pattern == "speeder"
longstr  <- cier_longstring(sim$responses)$flagged
pagetime <- cier_page_time(sim$page_seconds, sim$items_per_page)$flagged

c(speeders          = sum(speeder),
  caught_by_content = sum(longstr[speeder]),
  caught_by_pages   = sum(pagetime[speeder], na.rm = TRUE))
```

The content index flags none of them; per-page timing catches several. This is
why timing is converging evidence the screen cannot supply on its own -- and why
a real screen should pair content indices with the platform's timestamps.

## Method comparison: which index catches which pattern

To see each index on its home turf, plant *one* pattern at a time and score the
matched index. This reuses the `rank_auc()` helper above:

```{r}
one_pattern <- function(pattern, seed) {
  weights <- stats::setNames(rep(0, 8L),
    c("random", "straightline", "midpoint", "extreme",
      "diagonal", "alternating", "markov", "speeder"))
  weights[pattern] <- 1
  cier_simulate(n = 200L, items, prevalence = 0.25,
                patterns = weights, seed = seed)
}

cases <- data.frame(
  pattern = c("straightline", "alternating", "random", "diagonal", "speeder"),
  index   = c("longstring", "autocorrelation", "mahalanobis", "lazr",
              "total_time"),
  seed    = 11:15
)
cases$hi <- vapply(paste0("cier_", cases$index), flag_is_upper, logical(1L))

score_one <- function(sim, index, hi) {
  value <- switch(index,
    longstring      = cier_longstring(sim$responses)$value,
    autocorrelation = cier_autocorrelation(sim$responses, max_lag = 8L)$value,
    mahalanobis     = cier_mahalanobis(sim$responses)$value,
    lazr            = cier_lazr(sim$responses)$value,
    total_time      = cier_total_time(sim$seconds)$value)
  rank_auc(value, sim$truth$careless, higher_is_careless = hi)
}

cases$auc <- mapply(function(p, i, s, h) score_one(one_pattern(p, s), i, h),
                    cases$pattern, cases$index, cases$seed, cases$hi)
cases[, c("pattern", "index", "auc")]
```

The pattern-matched indices recover their target pattern almost perfectly, and
the `speeder` row again shows timing carrying the signal that content cannot --
score that same speeder data with a content index and the AUC drops to chance:

```{r}
speed_sim <- one_pattern("speeder", 15L)
c(timing  = rank_auc(cier_total_time(speed_sim$seconds)$value,
                     speed_sim$truth$careless, flag_is_upper("cier_total_time")),
  content = rank_auc(cier_longstring(speed_sim$responses)$value,
                     speed_sim$truth$careless, flag_is_upper("cier_longstring")))
```

One caveat for fair method comparison: by default every careless row also speeds
(`prop_speeded = 1`), so timing's edge over the content indices is partly built
in. Lower `prop_speeded` to leave that share of the *content*-careless rows at
attentive pace -- "slow careless" respondents who answer carelessly but
unhurriedly, whom timing should now miss:

```{r}
slow <- cier_simulate(n = 200L, items = items, prevalence = 0.25,
                      patterns = c(straightline = 1), prop_speeded = 0,
                      seed = 17L)
mean(slow$truth$speeded)   # 0: careless content, but attentive response times
```

No single index is a detector. The lesson the simulator makes visible is the
same one the screen is built on: different indices specialize in different
patterns, so detection comes from running them together and reading their
agreement.

## Power analysis with replications

Repeating a simulate-and-score run over many seeds turns a single accuracy
number into a sampling distribution. A short loop keeps this vignette inside the
build's timing budget; in practice you would run hundreds, possibly under
`future::plan(multisession)`, and may restrict `methods =` to the indices you
care about:

```{r}
replicate_one <- function(seed) {
  s <- cier_simulate(n = 150L, items, prevalence = 0.2,
                     patterns = c(random = 0.5, straightline = 0.5),
                     seed = seed)
  scr <- cier_screen(s$responses, s$items,
    control = list(cier_personal_reliability = list(seed = 1),
                   cier_gnormed = list(seed = 1)))
  pred <- seq_len(nrow(s$responses)) %in% cier_flagged_cases(scr, min_votes = 2)
  c(sensitivity = sum(s$truth$careless & pred) / max(sum(s$truth$careless), 1L),
    specificity = sum(!s$truth$careless & !pred) /
      max(sum(!s$truth$careless), 1L))
}

res <- vapply(101:105, replicate_one, numeric(2))
round(res, 3)
round(rowMeans(res), 3)
```

Sweep `prevalence`, `n`, the `patterns` mix, or `fpr` (through `control`) across
such a loop to map the operating characteristics of a design before collecting
any data.

## What simulation can and cannot tell you

The simulator is honest about its own limits, and so should its user be:

- **It can** rank methods on a controlled pattern, localize where each index's
  power comes from, size a study, and reproduce a published design. The patterns
  here are clean, so pattern-matched indices look strong -- *on these patterns*.
- **It cannot** stand in for real-world accuracy. Real careless responding is
  messier and less separable than any planted pattern; Schroeders et al. (2022)
  showed a simulation-trained advantage failing to transfer. Keep every
  real-world accuracy claim anchored to labelled data
  (`vignette("published-results")`), and report simulated results in their own
  clearly-labelled section -- never folded into a real-data league table.

Used that way -- as a design and comparison tool, not a validity claim -- the
simulator answers questions no labelled dataset can, because only here is the
truth known exactly.

## References

Bowling, N. A., Huang, J. L., Brower, C. K., & Bragg, C. B. (2023). The quick
and the careless: The construct validity of page time as a measure of
insufficient effort responding to surveys. *Organizational Research Methods,
26*(2), 323-352.

Curran, P. G. (2016). Methods for the detection of carelessly invalid responses
in survey data. *Journal of Experimental Social Psychology, 66*, 4-19.

Goldammer, P., Stöckli, P. L., Escher, Y. A., Annen, H., Jonas, K., & Antonakis,
J. (2024). Careless responding detection revisited: Accuracy of direct and
indirect measures. *Behavior Research Methods, 56*(8), 8422-8449.

Schroeders, U., Schmidt, C., & Gnambs, T. (2022). Detecting careless responding
in survey data using stochastic gradient boosting. *Educational and
Psychological Measurement, 82*(1), 29-56.

Welz, M., & Alfons, A. (2023). When respondents don't care anymore: Identifying
the onset of careless responding. *arXiv:2303.07167*.
